Reducing Acceptance Marks\\ in Emerson-Lei Automata by QBF-solving

### 2022-02-22 paper/main.tex
progress in intro

\title{Reducing Acceptance Marks\\ in Emerson-Lei Automata by QBF-solving}
\author{Tereza Schwarzov\'{a}}{Faculty of Informatics, Masaryk University, Brno, Czech Republic}{xschwar3@mail.muni.cz}{}{}
\author{Jan Strej\v{c}ek}{Faculty of Informatics, Masaryk University, Brno, Czech Republic}{strejcek@fi.muni.cz}{https://orcid.org/0000-0001-5873-403X}{}
\author{Juraj Major}{Faculty of Informatics, Masaryk University, Brno, Czech Republic}{major@fi.muni.cz}{}{}
\authorrunning{T. Schwarzov\'{a}, J. Strej\v{c}ek, and J. Major}
\Copyright{Tereza Schwarzov\'{a}, Jan Strej\v{c}ek, and Juraj Major}
\ccsdesc[500]{Theory of computation~Logic}
\ccsdesc[500]{Theory of computation~Automata over infinite objects}
\keywords{Emerson-Lei automata, TELA, automata reduction, QBF, \todo{telatko?}}
\funding{This work has been supported by the Czech Science Foundation grant GA19-24397S.}

@@

\begin{abstract}
  \emph{Transition-based Emerson-Lei automata (TELA)} become a popular
  formalism that generalizes many traditional kinds of automata over
  infinite words including Büchi, co-Büchi, Rabin, Streett, and parity
  automata. Transitions in a TELA are labelled with acceptance marks
  and its accepting formula is a positive Boolean combination of terms
  saying that a particular mark has to be visited infinitely or
  finitely often in an accepting run. Algorithms processing these
  automata are often very sensitive to the number of different
  acceptance marks. We introduce a new technique reducing the number
  of different acceptance marks in a TELA with use of \emph{quantified
    Boolean formulas (QBF)}: if a constructed QBF formula is
  satisfiable then its model encodes an equivalent accepting formula
  with less acceptance marks. In fact, we present three formula
  constructions trading the speed of the technique for higher
  reduction efficiency. The technique is implemented and evaluated on
  TELA produced by state-of-the-art tools Delag, ltl3tela,
  Rabinizer~4, and SPOT. The technique reduced some acceptance marks
  in automata produced by all these tools. In the case of Rabinizer~4,
  the technique reduced the cummulative number of acceptance marks to
  less than one half. 
\end{abstract}

@@

In 1987, Emerson and Lei~\cite{emerson.87.scp} introduced automata
over infinite words where acceptance conditions are arbitrary
combinations of acceptance primitives saying that a certain set of
states should be visited finitely often or infinitely often. In 2015,
the same kind of acceptance condition was described in the \emph{Hanoi
  omega-automata format (HOAF)}~\cite{babiak.15.cav}. The only
difference is that the acceptance primitives talk about finitely or
infinitely often visited acceptance marks rather than sets of states.
Acceptance marks are placed on transitions and each mark identifies
the set of transitions containing this mark. Hence, these automata are
called \emph{transition-based Emerson-Lei automata (TELA)} and they
generalize many traditional kinds of automata over infinite words
including Büchi, co-Büchi, Rabin, Streett, Muller or parity
automata.

TELA have attracted a lot of attention during the last five years.
Their popularity comes probably from the fact that these automata can
often use less states than equivalent automata with simpler acceptance
conditions. Further, algorithms handling TELA can automatically handle
all automata with traditional acceptance conditions. TELA automata can
be obtained for example by translating LTL formulas by
Delag~\cite{muller.17.gandalf}, ltl3tela~\cite{major.19.atva},
Rabinizer~4~\cite{kretinsky.18.cav}, or
SPOT~\cite{duret.16.atva2}. There are also some algorithms processing
these automata, for example the emptiness check~\cite{baier.19.atva}
or translation of TELA to parity automata~\cite{renkin.20.atva}.

\todo{Nechcem do experimentu pridat Delag?}

@@

@@

We have implemented both reduction techniques in a tool called
\texttt{telatko}. We show that the tool can reduce acceptance marks in
automata produced by ltl3tela~\cite{major.19.atva},
Rabinizer~4~\cite{kretinsky.18.cav}, and SPOT~\cite{duret.16.atva2}.
While the reduction is relatively modest for TELA produced by ltl3tela
and SPOT, it is substantial for automata produced by Rabinizer~4.

@@

>>>>

\title{Reducing Acceptance Marks\\ in Emerson-Lei Automata by QBF-solving}
\author{Tereza Schwarzov\'{a}}{Faculty of Informatics, Masaryk University, Brno, Czech Republic}{xschwar3@mail.muni.cz}{}{}
\author{Jan Strej\v{c}ek}{Faculty of Informatics, Masaryk University, Brno, Czech Republic}{strejcek@fi.muni.cz}{https://orcid.org/0000-0001-5873-403X}{}
\author{Juraj Major}{Faculty of Informatics, Masaryk University, Brno, Czech Republic}{major@fi.muni.cz}{}{}
\authorrunning{T. Schwarzov\'{a}, J. Strej\v{c}ek, and J. Major}
\Copyright{Tereza Schwarzov\'{a}, Jan Strej\v{c}ek, and Juraj Major}
\ccsdesc[500]{Theory of computation~Logic}
\ccsdesc[500]{Theory of computation~Automata over infinite objects}
\keywords{Emerson-Lei automata, TELA, automata reduction, QBF, telatko}
\funding{This work has been supported by the Czech Science Foundation grant GA19-24397S.}

@@

\begin{abstract}
  \emph{Transition-based Emerson-Lei automata (TELA)} become a popular
  formalism that generalizes many traditional kinds of automata over
  infinite words including Büchi, co-Büchi, Rabin, Streett, and parity
  automata. Transitions in a TELA are labelled with acceptance marks
  and its accepting formula is a positive Boolean combination of terms
  saying that a particular mark has to be visited infinitely or
  finitely often. % in an accepting run.
  Algorithms processing these automata are often very sensitive to the
  number of %different
  acceptance marks. We introduce a new technique reducing the number
  of %different
  acceptance marks in TELA with use of \emph{quantified Boolean
    formulas (QBF)}: if a constructed QBF formula is satisfiable then
  its model encodes an equivalent accepting formula with less
  acceptance marks. In fact, we present three formula constructions
  trading the speed of the technique for higher reduction
  efficiency. We evaluated our reduction technique on TELA produced by
  state-of-the-art tools Delag, ltl3tela, Rabinizer~4, and SPOT. The
  technique reduced some acceptance marks in automata produced by all
  these tools. In the case of Rabinizer~4, the cumulative number of
  acceptance marks was reduced to less than one half.
\end{abstract}

@@

In 1987, Emerson and Lei~\cite{emerson.87.scp} introduced automata
over infinite words where acceptance conditions are arbitrary
combinations of acceptance primitives saying that a certain set of
states should be visited finitely often or infinitely often. In 2015,
the same kind of acceptance condition was described in the \emph{Hanoi
  omega-automata format (HOAF)}~\cite{babiak.15.cav}. The only
difference is that the acceptance primitives talk about finitely or
infinitely often visited acceptance marks rather than sets of states.
Acceptance marks are placed on transitions and each mark identifies
the set of transitions containing this mark. Hence, these automata are
called \emph{transition-based Emerson-Lei automata (TELA)} and they
generalize many traditional kinds of automata over infinite words
including Büchi, co-Büchi, Rabin, Streett, and parity automata.

TELA have attracted a lot of attention during the last five years.
Their popularity comes probably from the fact that these automata can
often use less states than equivalent automata with simpler acceptance
conditions. Further, algorithms handling TELA can automatically handle
all automata with traditional acceptance conditions. TELA automata can
be obtained for example by translating LTL formulas by
Delag~\cite{muller.17.gandalf}, ltl3tela~\cite{major.19.atva},
Rabinizer~4~\cite{kretinsky.18.cav}, or
SPOT~\cite{duret.16.atva2}. There are also some algorithms processing
these automata, for example the emptiness check~\cite{baier.19.atva}
or translation of TELA to parity automata~\cite{renkin.20.atva}. \todo{Doplnit!}

@@

\todo{Honza je tu!}

@@

We have implemented both reduction techniques in a tool called
\texttt{telatko}. We show that the tool can reduce acceptance marks in
automata produced by Delag~\cite{muller.17.gandalf},
ltl3tela~\cite{major.19.atva}, Rabinizer~4~\cite{kretinsky.18.cav},
and SPOT~\cite{duret.16.atva2}.  While the reduction is relatively
modest for TELA produced by ltl3tela and SPOT, it is substantial for
automata produced by Rabinizer~4.

@@

There are some SAT-based approaches aimed to reduce the number of
states of automata over infinite words. Namely, there are reductions
designed for nondeterministic Büchi automata~\cite{ehlers.10.spin},
deterministic Büchi automata~\cite{ehlers.10.sat}, and deterministic
generalized Büchi automata~\cite{baarir.14.forte}. Further, there is
also a SAT-based approach that transforms a deterministic TELA
automaton to an equivalent automaton with a given acceptance condition
and a given number of states~\cite{baarir.15.lpar} (if such an
automaton exists). Note that all these techniques are usually very
slow and their authors typically suggest to use them only for specific
purposes like looking for cases where some automata construction can
be improved.
